---
title: "KPZ314 Camera-trap Practical"
format: html
execute:
  echo: true
---

```{r setup, include=FALSE}
targets::tar_config_set(script = "_targets.R")
targets::tar_source("R")

# auto‑build if the cache is empty
if (!targets::tar_exist_objects(names = "raw")) {
  message("Building targets (~3‑5 min on fresh clone)…")
  targets::tar_make()
}

# Build any missing targets needed for detectability/sensitivity sections
needed <- c(
  "psi_curves_all","fig_psi_curves","fig_det_heat",
  "tab_occu","fig_occu",
  "alpha_delta_unknown_tbl","beta_delta_unknown"
)
missing <- needed[!targets::tar_exist_objects(names = needed)]
if (length(missing)) {
  message("Building missing targets: ", paste(missing, collapse = ", "))
  targets::tar_make(names = missing)
}

raw            <- targets::tar_read(raw)
site           <- targets::tar_read(site)
taxa           <- targets::tar_read(taxa)
alpha          <- targets::tar_read(alpha)
comm           <- targets::tar_read(comm)
beta_bc        <- targets::tar_read(beta_bc)
beta_jac       <- targets::tar_read(beta_jac)
fig_heat       <- targets::tar_read(fig_heat)
fig_heat_bc    <- targets::tar_read(fig_heat_bc)
fig_heat_jac   <- targets::tar_read(fig_heat_jac)
fig_nmds       <- targets::tar_read(fig_nmds)
fig_nmds_jac   <- targets::tar_read(fig_nmds_jac)
nmds_stats     <- targets::tar_read(nmds_stats)
fig_site_map   <- targets::tar_read(fig_site_map)
fig_gap        <- targets::tar_read(fig_gap)
gamma_all      <- targets::tar_read(gamma_all)
gamma_region   <- targets::tar_read(gamma_region)
rare_region    <- targets::tar_read(rare_region)
rare_raw       <- targets::tar_read(rare_raw)
beta_permanova <- targets::tar_read(beta_permanova)
fig_sp_trends  <- targets::tar_read(fig_sp_trends)
rai            <- targets::tar_read(rai)
tab_effort     <- targets::tar_read(tab_effort)
alpha_ci       <- targets::tar_read(alpha_ci)
gamma_by_habitat <- targets::tar_read(gamma_by_habitat)
gamma_all_ci     <- targets::tar_read(gamma_all_ci)

# GLM module objects
glm_diag_all     <- targets::tar_read(glm_diag_all)
glm_irrs_all     <- targets::tar_read(glm_irrs_all)
fig_glm_coef     <- targets::tar_read(fig_glm_coef)
fig_glm_pd_type  <- targets::tar_read(fig_glm_pd_type)
```

## 1. Orientation

**Look at the raw, expert‑checked data table.**

```{r}
dplyr::glimpse(raw)
```

> **TASK 1.0** Think about why the data are encoded in this way.

### 1.1 Map of camera sites

```{r}
fig_site_map
```

> **TASK 1.1** The sites are clearly separated spatially. Is this the only relevant difference?

---

## 2. Independent events and effort

We treat “independent events” as detections separated by at least a configurable gap. Below, explore how site‑level metrics change across gap values (1, 5, 10, 30 minutes).

```{r}
fig_gap
```

Sampling effort affects interpretation, so we compute RAIs per 100 trap‑nights and provide the effort manifest for transparency.

```{r}
tab_effort
```

```{r}
head(rai, 8)
```

> **TASK 2.1** Pick a gap that balances independence and sample size for this fauna. Justify briefly (≤ 40 words).

```{r}
# Any sites with zero total events (dropped from turnover distances)
comm_mat <- targets::tar_read(comm)
zero_sites <- names(which(rowSums(comm_mat) == 0))
if (length(zero_sites)) {
  cat("Note: Dropped zero-event sites from turnover distances: ", paste(zero_sites, collapse = ", "), "\n")
}
```

---

## 3. Site‑level diversity

**Inspect alpha‑diversity metrics for all sites.**

```{r}
# Offset counts by excluding ubiquitous 'unknown' class: subtract 1
alpha_adj <- dplyr::mutate(alpha,
  richness = pmax(richness - 1, 0),
  chao1    = pmax(chao1 - 1, 0)
)

# Panel 1: abundance-like metrics (counts)
alpha_long_counts <- tidyr::pivot_longer(
  dplyr::select(alpha_adj, site, richness, chao1),
  -site, names_to = "metric", values_to = "value"
)

ggplot2::ggplot(alpha_long_counts, ggplot2::aes(x = metric, y = value)) +
  ggplot2::geom_boxplot()
```

```{r}
# Panel 2: entropy/evenness metrics
alpha_long_entropy <- tidyr::pivot_longer(
  dplyr::select(alpha_adj, site, pielou, shannon, simpson),
  -site, names_to = "metric", values_to = "value"
)

ggplot2::ggplot(alpha_long_entropy, ggplot2::aes(x = metric, y = value)) +
  ggplot2::geom_boxplot()
```

Hill numbers provide a unified view of diversity: q=0 is richness, q=1 is the effective number of species from Shannon (exp[H]), and q=2 is the inverse Simpson (dominance‑sensitive).

Note: q=0 counts species (richness), q=1 weights by common species, q=2 emphasises dominants; divergence between q=0 and q=2 signals evenness differences. In our tables, hill_q0 ≡ richness.

```{r}
# Bootstrap 95% CIs for Hill numbers and Pielou (per site)
knitr::kable(head(alpha_ci, 12), caption = "Alpha diversity (Hill) with percentile bootstrap 95% CIs (first 12 rows)")
```

```{r}
# Sites with zero events (excluded from alpha CI bootstrap)
 ev <- targets::tar_read(events_default)
 site_counts <- dplyr::count(ev, site, name = "events")
 zero_alpha <- setdiff(unique(ev$site), site_counts$site)
 if (length(zero_alpha)) {
   cat("Note: Sites excluded from alpha bootstrap (zero events): ", paste(zero_alpha, collapse = ", "))
 }
```

> **TASK 3.1** Choose one site whose Shannon index is notably high or low. State an ecological hypothesis to explain the pattern (≤ 40 words).

### 3.2  Total (γ) diversity

```{r}
# Subtract 1 to discount the 'unknown' species placeholder
gamma_all_adj    <- gamma_all - 1L
gamma_region_adj <- dplyr::mutate(gamma_region, gamma = gamma - 1L)

gamma_all_adj     # scalar (adjusted)
gamma_region_adj  # by type (adjusted)

# Habitat-level gamma (Hill numbers) with 95% CIs
knitr::kable(gamma_by_habitat, caption = "Habitat-level gamma diversity (Hill) with site-bootstrap 95% CIs")
```

Seeds: The pipeline sets a global seed via tar_option_set(seed = 1L); alpha and gamma bootstraps also pass seed = 1L for reproducibility.

> Note: γ CIs are from site‑level resampling of pooled counts, reflecting among‑site heterogeneity rather than event noise.
> **TASK 3.2** How much of the γ diversity is captured in wet vs dry sites? (≤ 30 words)

### 3.3  Sampling sufficiency

```{r}
rare_raw
```

> **TASK 3.3.1** Are some of the camera sites undersampled? (≤ 30 words)

```{r}
rare_region
```

Sampling sufficiency and coverage: the endpoints on each rarefaction curve show the observed richness and its estimated coverage (S_obs/Chao1). Closer to 1 implies better sampling completeness.

> **TASK 3.3.2** Do the wet and dry curves suggest similar sampling completeness across habitats? Explain in ≤ 40 words.

> **TASK 3.3.3** Report the coverage values at the endpoints for wet vs dry. Which habitat is closer to asymptote?

---

## 4. Community turnover

Bray–Curtis (abundance) and Jaccard (presence–absence) emphasise different aspects of community turnover. Report results for both: abundance patterns (dominance) vs composition (which species are present).

### 4.1  Within-habitat dissimilarities and community structure

```{r}
fig_heat_bc
```

```{r}
fig_heat_jac
```

```{r}
cat(turnover_caption())
```

```{r}
plotly::ggplotly(fig_nmds)
```

```{r}
fig_nmds_jac
```

```{r}
knitr::kable(nmds_stats, caption = "NMDS fit statistics (stress) for each distance")
```

> Note: If stress > 0.2, interpret NMDS geometry cautiously; high stress indicates a poor low‑dimensional fit.

Guidance:
- If disp_p ≥ 0.05 and p_perm < 0.05: report location separation with R² and p; dispersion is homogeneous.
- If disp_p < 0.05: “Groups differ in dispersion; R² mixes location and spread; check medians and NMDS cautiously.”
- Metric choice: Jaccard asks “different species present?”; Bray–Curtis asks “different composition and/or dominance?”. Report both.


> **TASK 4.1** Name two clusters of sites that appear compositionally similar. Suggest one abiotic driver that could explain the pattern (≤ 40 words).

### 4.2  Are wet vs dry communities distinct (location) and equally dispersed?

```{r}
beta_permanova
```

Permutation seeds for Bray and Jaccard are matched (seed = 1) so p-value differences reflect the metric, not RNG.

Report R² (effect size) and permutation p for each distance. If `dispersion p < 0.05`, then groups differ in spread; interpret R² as partly/mostly driven by dispersion rather than centroid separation (location).

Hulls show convex envelopes, not variance. Use `disp_p` to assess spread differences.

```{r}
# Effect contrast: ΔR² = R²_bray − R²_jaccard
targets::tar_read(beta_r2_delta)
```

```{r}
# Optional quick sanity check: median within vs between distances
wb_bray <- within_between_summary(beta_bc, site, "type")
wb_jac  <- within_between_summary(beta_jac, site, "type")
knitr::kable(dplyr::bind_rows(wb_bray, wb_jac), caption = "Median within- vs between-habitat distances")
```

---

## 5. Habitat selection (effort‑offset GLMs)

We fit effort‑offset GLMs for five focal taxa with sensible family selection (Poisson ⇒ quasi/NB if overdispersed). Numeric covariates are z‑scored, so IRRs are per +1 SD. Models include a log(trap_nights) offset.

```{r}
knitr::kable(glm_diag_all, caption = "GLM diagnostics (family, dispersion, sites)")
```

```{r}
knitr::kable(glm_irrs_all, caption = "Incidence Rate Ratios (IRR) with 95% Wald CIs (quasi uses dispersion‑inflated SEs)")
```

If the diagnostics table notes “type dropped … separation”, inference on wet vs dry failed for that species because all events occurred in one habitat. This is a common issue for rare taxa; we drop `type` and proceed with other covariates.

```{r}
# one forest plot per species
for (p in fig_glm_coef) print(p)
```

```{r}
# partial dependence for wet vs dry (rates per 100 trap‑nights)
for (p in fig_glm_pd_type) print(p)
```

> TASK 5.1 Pick one species and interpret the `type` IRR (wet vs dry) with its 95% CI in ≤ 35 words.
>
> TASK 5.2 Which habitat covariate has the strongest effect (IRR farthest from 1)? Interpret it per +1 SD in ≤ 35 words.

---

## 6. Imperfect detectability (occupancy)

Even when a species is present, we don’t always detect it on a given day (p < 1). Below, we build daily detection histories, show how naïve occupancy ψ̂(t) rises with effort and differs by habitat, and fit a simple occupancy model for a common focal species (bennetts_wallaby).

### 6.1 Daily detection histories

```{r}
det_list <- targets::tar_read(fig_det_heat)

# Prefer a mid-frequency species (brushtail_possum) if available
printed <- FALSE
for (i in seq_along(det_list)) {
  sp <- attr(det_list[[i]], "species")
  if (!is.null(sp) && sp == "brushtail_possum") { print(det_list[[i]]); printed <- TRUE; break }
}
if (!printed && length(det_list)) print(det_list[[1]])
```

> Assumptions: days without detections are treated as non-detections; cameras are assumed operating continuously for `op_days`. We cap detection histories at up to 21 days (or fewer if deployments were shorter).
> Timezone: Days are computed in Hobart local time (Australia/Hobart) to avoid UTC/DST boundary surprises; only the calendar date is used when forming daily occasions.

### 6.2 Naïve occupancy ψ̂(t) by habitat

```{r}
psi_all <- targets::tar_read(psi_curves_all)
knitr::kable(dplyr::filter(psi_all, day %in% c(3,7,14)),
             caption = "Naïve occupancy ψ̂(t) at days 3, 7, and 14 (by habitat)")
```

> Denominators: All sites contribute through day `det_days_max`; we cap curves at that window so denominators remain constant when comparing ψ̂(t) across days and habitats.

```{r}
det_days_max <- targets::tar_read(det_days_max)
cat("Note: det_days_max =", det_days_max, "days\n")
```

```{r}
figs <- targets::tar_read(fig_psi_curves)
for (p in figs) print(p)
```

> TASK 6.1 For one species, how many days of effort are needed for ψ̂(t) to exceed 0.8 in wet vs dry sites?

### 6.3 Occupancy demo (auto-picked species)

```{r}
picked <- targets::tar_read(occ_species_auto)
cat("Demo species:", picked, "\n")

knitr::kable(targets::tar_read(tab_occu),
             caption = "Occupancy demo (ψ by habitat; p is per-day detection probability)")
```

```{r}
targets::tar_read(fig_occu)
```

> TASK 6.2 Report the overall ψ̂ with its 95% CI. What does the estimated per‑day detection probability p imply about required survey effort? (Use the ψ̂(t) curves by habitat for context.)

Note: We fit ψ ~ type, p ~ 1. If habitat groups are nearly separated or detections are extremely sparse, the likelihood can become flat and the Hessian singular; predictions on the probability scale can still be interpreted cautiously. If instability arises, use the ψ̂(t) curves to discuss differences.

> Guardrail: If ψ̂_wet ≠ ψ̂_dry but p is low, extra effort (more days) can change naïve conclusions. State this explicitly in your answer.

#### 6.3a Species picker rationale

We auto-pick a “pedagogical” species by choosing the one whose windowed detection rate is closest to 50% of sites (ideally within 20–80%). This yields enough 0s and 1s across sites for stable estimation and clear contrasts.

```{r}
if (!targets::tar_exist_objects(names = "det_detect_summary_all")) {
  targets::tar_make(names = c("det_histories","det_detect_summary","det_detect_summary_all"))
}

rank_tbl <- targets::tar_read(det_detect_summary_all) |>
  dplyr::mutate(closeness = abs(prop_overall - 0.5)) |>
  dplyr::arrange(closeness, dplyr::desc(prop_overall)) |>
  dplyr::mutate(
    prop_overall = round(prop_overall, 3),
    prop_dry     = round(prop_dry, 3),
    prop_wet     = round(prop_wet, 3),
    closeness    = round(closeness, 3)
  ) |>
  head(7)

knitr::kable(rank_tbl, caption = "Species ranked by windowed detection rate closeness to 0.5 (lower is better)")
```

### 6.4 Sensitivity: excluding ‘unknown_animal’

```{r}
if (!targets::tar_exist_objects(names = "alpha_delta_unknown_tbl")) {
  targets::tar_make(names = "alpha_delta_unknown_tbl")
}
knitr::kable(targets::tar_read(alpha_delta_unknown_tbl),
             caption = "Alpha metric deltas (no 'unknown_animal' minus all)")
```

```{r}
targets::tar_read(beta_delta_unknown)
```

```{r}
# Small effect-size summary students can cite
sens <- targets::tar_read(beta_delta_unknown)
knitr::kable(sens, caption = "Unknown-animal sensitivity: mean |Δ| and Pearson r for Bray/Jaccard (upper triangle)")
```

> Interpretation hint: If r is near 1 and mean |Δ| is small, distances barely move when excluding ‘unknown_animal’. If r is lower and mean |Δ| is non-trivial, composition/dominance shifts materially.

> TASK 6.3 Which metrics appear robust to excluding ‘unknown_animal’? Briefly justify.

---

### 4.3  Species spotlight

```{r}
fig_sp_trends
```

> **TASK 4.3** Which habitat records more possum events? Suggest one ecological driver (≤ 40 words).

---

## 6. Trait exploration (stretch goal)

**Link alpha metrics to species traits (from the taxa table).**

```{r}
# Example structure view (joined alpha + site metadata)
alpha_site <- dplyr::left_join(alpha, site, by = "site")
utils::str(alpha_site)
```

> **TASK 5.1** If larger‑bodied taxa are more detectable, what pattern would you expect in the alpha metrics? Test it with a quick plot or summary, and comment (≤ 50 words).

---

## 7. Reflection

> **TASK 6.1** In ≤ 50 words, reflect on one limitation or source of bias in using camera‑trap data for community‑level inference in Tasmanian fauna.

---

## 8. Free Exploration

You can load any pipeline object into your R environment and explore it.

**For example:**

```{r}
# Assign the 'alpha' site diversity table to your workspace
alpha_local <- targets::tar_read(alpha)

# view the whole table in RStudio
alpha_local
```

**Try these ideas:**

* **Compare any two objects**
  e.g. plot Chao1 vs. Shannon diversity for all sites

  ```{r}
  alpha_tbl <- targets::tar_read(alpha)
  plot(alpha_tbl$chao1, alpha_tbl$shannon,
       xlab = "Chao1", ylab = "Shannon", pch = 16)
  ```

* **Investigate site metadata**

  ```{r}
  site_tbl <- targets::tar_read(site)
  summary(dplyr::select(site_tbl, type, leaf_litter, cwd, can_foliage))
  ```

* **Make a custom plot**

  ```{r}
  taxa_tbl <- targets::tar_read(taxa)
  ggplot2::ggplot(dplyr::filter(taxa_tbl, !is.na(status) & status != "na"),
                  ggplot2::aes(x = status, fill = status)) +
    ggplot2::geom_bar(show.legend = FALSE) +
    ggplot2::labs(title = "Taxon status in Tasmanian fauna",
        x = "Status", y = "Frequency") +
    ggplot2::theme_minimal()
  ```

> **TASK 7.1** Are native or introduced taxa more common in this camera-trap dataset? What does this suggest about Tasmania’s fauna?

* **Re-run a helper function yourself**

  Try feeding a subset of data to `calc_alpha()` or `build_comm_matrix()`.

  ```{r}
  # For interactive use of helpers:
  source("R/helpers_metrics.R")

  joined <- targets::tar_read(joined)
  custom <- joined[joined$type == "dry", ]
  calc_alpha(custom)
  ```

**Or try anything else**: every object from the workflow is just a data frame or matrix you can inspect, plot, or save!

---

> **Bonus:** In the R console, run `tar_manifest()` to see all available pipeline objects, or use `ls()` to see what’s loaded into your workspace.

---

## 9. Your Own Analysis

Write a short script or chunk below where you try something new with the data—anything goes!

```{r}

# Your code here!

```

---

## 9. “In-console” play

After knitting, you can open the R console, load the project, and call `tar_read()` for any target, or `source("R/helpers_metrics.R")` to get all helpers.

For truly unchained analysis: create an `explore.R` or `scratch.Rmd` and go free-form!
